{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-03 thesis/overview.tex
use cleveref

Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
},
but whether this constitutes a bug is a matter of perspective.
Was the program written to count to 2000, or was it written to compute a randomized distribution?
In this thesis, we make no attempt to reason about the ``intent'' of programs,
so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
according to commonly-accepted notions of \revisionminor{which} programs behaviours are always bad.
%
Bug conditions include assertion failures,
memory access errors (i.e., segmentation fault or bus error),
heap errors (i.e., use-after-free or overflow),
deadlocks,
and infinite loops (which must be identified heuristically \cite{entscheidungsproblem}).

@@

To prevent unexpected interleavings such as the example in Figure~\ref{fig:concurrency-bug}(b),
most concurrent programs use {\em concurrency primitives} to control which interleavings are possible.
Controlling nondeterminism is not typically provided by any features of programming languages themselves;
rather, it is achieved via special atomicity mechanisms provided by the CPU and/or operating system -- hence the term ``primitive''.
For example, x86 CPUs provide the {\tt xchg} instruction, which performs both a read and subsequent write to some shared memory, with no possibility for other logic to interleave in between.
Using such atomic instructions as building blocks, concurrency libraries provide abstractions for controlling nondeterminism in several commonly-desired ways.
These include {\em locks}, {\em descheduling}, {\em condition variables}, {\em semaphores}, {\em reader-writer locks}, and {\em message-passing}.

\begin{figure}[t]
	\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
		{\small
		\begin{tabular}{l}
			\texttt{\ctype{typedef struct mutex} \{} \\
			\texttt{~~~~\ctype{volatile int} held;} \\
			%\texttt{~~~~\ctype{int} owner;} \\
			\texttt{\} \ctype{mutex\_t};} \\
			\texttt{\ctype{void} \call{mutex\_lock}(\ctype{mutex\_t} *mp) \{}\\
			\texttt{~~~~\flow{while} (\call{xchg}(mp->held, \const{1}))} \\
			%\texttt{~~~~~~~~\call{yield}(mp->owner);} \\
			\texttt{~~~~~~~~\call{yield}(-1);} \\
			%\texttt{~~~~mp->owner = \call{gettid}();} \\
			\texttt{\}} \\
			\texttt{\ctype{void} \call{mutex\_unlock}(\ctype{mutex\_t} *mp) \{}\\
			%\texttt{~~~~mp->owner = \ctype{-1};} \\
			\texttt{~~~~mp->held = \ctype{0};} \\
			\texttt{\}} \\
		\end{tabular}
		}
		&
		{\small
		\begin{tabular}{l}
			\texttt{\ctype{int} x;} \\
			\texttt{\ctype{mutex\_t} m;} \\
			\texttt{\ctype{void} \call{count}() \{} \\
			\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
			\texttt{~~~~~~~~\call{mutex\_lock}(\&m);} \\
			\texttt{~~~~~~~~x++;} \\
			\texttt{~~~~~~~~\call{mutex\_unlock}(\&m);} \\
			\texttt{~~~~\}} \\
			\texttt{\}} \\
		\end{tabular}
		}
		\\
		(a) A simple mutual exclusion lock built using the {\tt xchg} instruction. %(accessed using a GCC compiler intrinsic).
		&
		(b) The {\tt count} function from Figure~\ref{fig:concurrency-bug}, adjusted to use a mutex to ensure each increment of {\tt x} is uninterruptible.
	\end{tabular}
	\caption[Using a locking primitive to protect accesses to shared state.]
		{Using a locking primitive to protect accesses to shared state.}
	\label{fig:mutex}
\end{figure}

Each such abstraction provides certain semantics about \revisionminor{which} thread interleavings can arise surrounding their use.
When building a tool for testing concurrent programs,
one may include some computational understanding of the behaviour of any, or all, of these abstractions.
Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
and allows the testing tool to reduce the possible space of interleavings
(or the set of false positive data-race candidates reported, etc.;
\revisionminor{see \sect{\ref{sec:landslide-config-landslide}}})
at the cost of increasing the implementation and theoretical complexity of the analysis
\revisionminor{(\sect{\ref{sec:tm-abstraction}})}.
This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).

Locks (or {\em mutexes}, short for ``mutual exclusion locks'') are objects, shared by multiple threads, which allow the programmer to mark certain {\em critical sections} of code that must not interleave with each other.
When one thread completes a call to {\tt mutex\_lock(mp)}, all invocations by other threads on the same {\tt mp}
will \revisionminor{block (i.e., wait)} until the corresponding {\tt mutex\_unlock(mp)}.
Figure~\ref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
and (b) shows how a mutex may be used to fix the example from Figure~\ref{fig:concurrency-bug}.

@@

Transactional memory (TM) may be implemented either in hardware,
using special instructions and existing cache logic, %coherence algorithms,
or in software, via library calls and a log-based commit approach.
Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
Hardware transactions (HTM) \cite{intel-tsx-overview}
%\cite{htm-experience, htm-performance}
achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
The example program in Figure~\ref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
	The solution presented here is actually incomplete; stay tuned until Chapter~\ref{chap:tm} for the surprising twist!
},
using GCC's compiler intrinsics \cite{htm-gcc}.

\begin{figure}[t]
	\begin{center}
		\small
		\begin{tabular}{l}
		\texttt{\ctype{void} \call{count}() \{} \\
		\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
		\texttt{~~~~~~~~\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{\_xend}();} \\
		\texttt{~~~~~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_lock}(\&m);} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_unlock}(\&m);} \\
		\texttt{~~~~~~~~\}} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption[The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.]
		{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
		If the transaction in the top branch aborts,
		whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

Concerning possible execution patterns,
the main difference between STM and HTM is the circumstances under which a transaction may abort.
A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
Chapter~\ref{chap:tm} will explore the consequences of this difference further.
%
This thesis will focus on HTM as my platform for testing transactional programs,
to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.

@@

To understand what it means to exhaustively test all possible thread interleavings,
one must define the possible execution sequences as a finite {\em state space}.
To visualize this,
using a single iteration of the {\tt x++;} loop from Figure~\ref{fig:concurrency-bug} as an example,
with {\tt x++;} expanded into its three corresponding pseudo-assembly instructions,
Figure~\ref{fig:tree}(a) shows all possible execution interleavings
%of the compiled code
between 2 threads.

\begin{figure}[p]
	\begin{tabular}{c}
		\includegraphics[width=\textwidth]{statespace-list.pdf}
		\\
		(a) Interleavings visualized individually, as a list.
		\\
		\\
		\includegraphics[width=\textwidth]{statespace-tree.pdf}
		\\
		(b) Interleavings (same order as in (a)), with common prefixes \\
		combined as ``preemption points'', forming a tree.
	\end{tabular}
	\caption[Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.]
		{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
	Thread~1 is represented by purple ovals, thread~2 by yellow squares, and time flows from top to bottom.
	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
	the full state space is twice the size, and the other half is symmetric to the one shown.}
	\label{fig:tree}
\end{figure}

@@

During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
where interrupting the current thread to run a different one is likely to produce different behaviour.
These so-called {\em preemption points} may be identified by any combination of human intuition and machine analysis.
Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
Considering that at each preemption point multiple threads exist as options to run next,
the set of possible ways to execute the program can be viewed as a tree.
Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program,
using each pseudo-assembly instruction as a preemption point.

The number of preemption points in each execution defines the depth of this tree,
and the number of threads available to run defines the branching factor.
Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
Addressing the scaling problem in this exponential relation is the central research problem for all model checkers,
\revisionminor{which this thesis will address in Chapter~\ref{chap:quicksand}}.

@@

{\bf Concretely speaking:}
Figure~\ref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
The specifics of execution before the thread 1/thread 2 sequence,
other possible threads to run instead of threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
\sect{\ref{sec:landslide-dpor}} explains how DPOR implements such a deduction in more detail.

@@

Iterative Context Bounding \cite{chess-icb} is a popular such technique which heuristically reorders the search to prioritize interleavings with fewer preemptions first.
This heuristic is based on the insight that most bugs require few preemptions to uncover, so interleavings with a number of preemptions that exceeds a certain bound will be de-prioritized, only tested until after all the fewer-preemption interleavings are completed.
Preemption sealing \cite{sealing} is another heuristic strategy which restricts the scope of the search by limiting the model checker to use only preemption points arising from certain functions in the source code.
This allows developers to vastly reduce state space size by identifying which program modules are already trusted,
although it requires some human intuition to correctly mark those boundaries.
Iterative Deepening, presented in Chapter~\ref{chap:quicksand}, is another such search heuristic.

@@

Two instructions are said to race if:
\begin{enumerate}
	\item they both access the same memory address,
	\item at least one is a write,
	\item the threads do not hold the same lock,
	\item and no synchronization enforces an order on the thread transitions
		(the {\em Happens-Before} \revisionminor{relation~\cite{lamport-clocks,hybriddatarace}}, described below).
\end{enumerate}
%In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.

@@

\begin{itemize}
	\item Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
as shown in the example program in Figure~\ref{fig:hb-example}(a).
We call such unreported access pairs {\em false negatives}.

\item
\cite{hybriddatarace} introduces the {\em Limited Happens-Before} relation,
which will report such potential races
by considering only blocking operations like {\tt cond\_wait} to enforce the order.
However, consider the similar program in Figure~\ref{fig:hb-example}(b),
in which the access pair ceases to exist in the alternate interleaving.
Limited Happens-Before will report all potential races, avoiding false negatives \cite{tsan},
but at the cost of necessarily reporting some such {\em false positives}.

\item
In recent work, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
extends Happens-Before to additionally report a subset of potential races while soundly avoiding false positives.
It tracks conflicting accesses in intervening
critical sections to determine whether lock events are unrelated to a potential race.
Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
although it can still miss true potential races in other cases.
\end{itemize}

Landslide implements both Happens-Before (henceforth referred to as {\em Pure Happens-Before} for clarity) and Limited Happens-Before.
Chapter~\ref{chap:quicksand} includes a comparison of the two approaches for the purpose of finding new preemption points for model checking.
%, we use the Limited Happens-Before relation for our analysis.
%The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
%my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
%Hence, I accept some overhead from false positives for the sake of more thorough testing.

@@

The Pebbles kernel architecture
is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and \revisionminor{Implementation} \cite{kspec,thrlib}.
In the course of a semester, students work on five programming assignments;
the first two are individual, and the remaining three are the products of two-person teams.
I will focus on the third and fourth of these, the thread library and kernel,
called ``P2'' and ``P3'' respectively (the project numbers start at 0).
The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}.

@@

\subsubsection{P3}
In P3, students implement a kernel which provides the same system calls shown in Table~\ref{tab:syscalls}, previously provided by the reference kernel.
Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
This requires less implementation complexity than the more featureful
\revisionminor{{\tt rfork} model, of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model, of Linux}.

@@

The 15-410 course staff is notoriously secretive about the nature of many concurrency bugs
students commonly encounter during P2 and P3.
This is driven by a desire to cause students to find, diagnose, and fix these bugs on their own during the projects,
rather than to be surprised by them afterwards during grading
\cite{de0u-2018}.
%
One such example is the {\tt paraguay} unit test distributed with P2 (\sect{\ref{sec:education-pebbles-tests}}),
which targets a subtle condition-variable bug.
The test uses the {\tt misbehave} system call to \revisionminor{amplify the probability of} a particular thread interleaving,
likely to expose the bug,
which is otherwise very unlikely to arise in normal execution.
The reference kernel specification \cite{kspec} does not define the {\tt misbehave} modes' behaviours,
as doing so would deprive students of the learning experience of discovering the interleaving in question on their own.
%
I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.

@@

		see \sect{\ref{sec:background-datarace}}.

@@

		(\sect{\ref{sec:quicksand-soundness}}).

@@

		See \sect{\ref{sec:landslide-dpor}}.

@@

		See \sect{\ref{sec:landslide-estimate}}.

@@

		See \sect{\ref{sec:overview-statespace}}; Figure~\ref{fig:tree}.

@@

		are both {\em heuristic} (\sect{\ref{sec:landslide-blocking}}),

@@

	\item {\bf ICB.} Short for ``Iterative Context Bounding''. See \sect{\ref{sec:landslide-icb}}.

@@

		See \sect{\ref{sec:quicksand-id}}.

@@

		Documented in Chapter \ref{chap:landslide}.

@@

		See \sect{\ref{sec:overview-mutex}}.

@@

		is kept secret to preserve the intended challenge of debugging (see \sect{\ref{sec:410-secrecy}}).}

@@

		See \sect{\ref{sec:pebbles}}.

@@

		The educational kernel architecture used in 15-410 at CMU. See \sect{\ref{sec:pebbles}}.}

@@

		as well as several other universities. See \sect{\ref{sec:overview-pintos}}.}

@@

		See \sect{\ref{sec:landslide-pps}}.

@@

		See Chapter~\ref{chap:quicksand}.

>>>>

Consider the program in \Cref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
},
but whether this constitutes a bug is a matter of perspective.
Was the program written to count to 2000, or was it written to compute a randomized distribution?
In this thesis, we make no attempt to reason about the ``intent'' of programs,
so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
according to commonly-accepted notions of \revisionminor{which} programs behaviours are always bad.
%
Bug conditions include assertion failures,
memory access errors (i.e., segmentation fault or bus error),
heap errors (i.e., use-after-free or overflow),
deadlocks,
and infinite loops (which must be identified heuristically \cite{entscheidungsproblem}).

@@

To prevent unexpected interleavings such as the example in \Cref{fig:concurrency-bug}(b),
most concurrent programs use {\em concurrency primitives} to control which interleavings are possible.
Controlling nondeterminism is not typically provided by any features of programming languages themselves;
rather, it is achieved via special atomicity mechanisms provided by the CPU and/or operating system -- hence the term ``primitive''.
For example, x86 CPUs provide the {\tt xchg} instruction, which performs both a read and subsequent write to some shared memory, with no possibility for other logic to interleave in between.
Using such atomic instructions as building blocks, concurrency libraries provide abstractions for controlling nondeterminism in several commonly-desired ways.
These include {\em locks}, {\em descheduling}, {\em condition variables}, {\em semaphores}, {\em reader-writer locks}, and {\em message-passing}.

\begin{figure}[t]
	\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
		{\small
		\begin{tabular}{l}
			\texttt{\ctype{typedef struct mutex} \{} \\
			\texttt{~~~~\ctype{volatile int} held;} \\
			%\texttt{~~~~\ctype{int} owner;} \\
			\texttt{\} \ctype{mutex\_t};} \\
			\texttt{\ctype{void} \call{mutex\_lock}(\ctype{mutex\_t} *mp) \{}\\
			\texttt{~~~~\flow{while} (\call{xchg}(mp->held, \const{1}))} \\
			%\texttt{~~~~~~~~\call{yield}(mp->owner);} \\
			\texttt{~~~~~~~~\call{yield}(-1);} \\
			%\texttt{~~~~mp->owner = \call{gettid}();} \\
			\texttt{\}} \\
			\texttt{\ctype{void} \call{mutex\_unlock}(\ctype{mutex\_t} *mp) \{}\\
			%\texttt{~~~~mp->owner = \ctype{-1};} \\
			\texttt{~~~~mp->held = \ctype{0};} \\
			\texttt{\}} \\
		\end{tabular}
		}
		&
		{\small
		\begin{tabular}{l}
			\texttt{\ctype{int} x;} \\
			\texttt{\ctype{mutex\_t} m;} \\
			\texttt{\ctype{void} \call{count}() \{} \\
			\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
			\texttt{~~~~~~~~\call{mutex\_lock}(\&m);} \\
			\texttt{~~~~~~~~x++;} \\
			\texttt{~~~~~~~~\call{mutex\_unlock}(\&m);} \\
			\texttt{~~~~\}} \\
			\texttt{\}} \\
		\end{tabular}
		}
		\\
		(a) A simple mutual exclusion lock built using the {\tt xchg} instruction. %(accessed using a GCC compiler intrinsic).
		&
		(b) The {\tt count} function from \Cref{fig:concurrency-bug}, adjusted to use a mutex to ensure each increment of {\tt x} is uninterruptible.
	\end{tabular}
	\caption[Using a locking primitive to protect accesses to shared state.]
		{Using a locking primitive to protect accesses to shared state.}
	\label{fig:mutex}
\end{figure}

Each such abstraction provides certain semantics about \revisionminor{which} thread interleavings can arise surrounding their use.
When building a tool for testing concurrent programs,
one may include some computational understanding of the behaviour of any, or all, of these abstractions.
Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
and allows the testing tool to reduce the possible space of interleavings
(or the set of false positive data-race candidates reported, etc.;
\revisionminor{see \cref{sec:landslide-config-landslide}})
at the cost of increasing the implementation and theoretical complexity of the analysis
\revisionminor{(\cref{sec:tm-abstraction})}.
This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).

Locks (or {\em mutexes}, short for ``mutual exclusion locks'') are objects, shared by multiple threads, which allow the programmer to mark certain {\em critical sections} of code that must not interleave with each other.
When one thread completes a call to {\tt mutex\_lock(mp)}, all invocations by other threads on the same {\tt mp}
will \revisionminor{block (i.e., wait)} until the corresponding {\tt mutex\_unlock(mp)}.
\Cref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
and (b) shows how a mutex may be used to fix the example from \Cref{fig:concurrency-bug}.

@@

Transactional memory (TM) may be implemented either in hardware,
using special instructions and existing cache logic, %coherence algorithms,
or in software, via library calls and a log-based commit approach.
Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
Hardware transactions (HTM) \cite{intel-tsx-overview}
%\cite{htm-experience, htm-performance}
achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
The example program in \Cref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
	The solution presented here is actually incomplete; stay tuned until \Cref{chap:tm} for the surprising twist!
},
using GCC's compiler intrinsics \cite{htm-gcc}.

\begin{figure}[t]
	\begin{center}
		\small
		\begin{tabular}{l}
		\texttt{\ctype{void} \call{count}() \{} \\
		\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
		\texttt{~~~~~~~~\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{\_xend}();} \\
		\texttt{~~~~~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_lock}(\&m);} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_unlock}(\&m);} \\
		\texttt{~~~~~~~~\}} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption[The example {\tt count} routine from \Cref{fig:mutex}, rewritten to use HTM.]
		{The example {\tt count} routine from \Cref{fig:mutex}, rewritten to use HTM.
		If the transaction in the top branch aborts,
		whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

Concerning possible execution patterns,
the main difference between STM and HTM is the circumstances under which a transaction may abort.
A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
\Cref{chap:tm} will explore the consequences of this difference further.
%
This thesis will focus on HTM as my platform for testing transactional programs,
to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.

@@

To understand what it means to exhaustively test all possible thread interleavings,
one must define the possible execution sequences as a finite {\em state space}.
To visualize this,
using a single iteration of the {\tt x++;} loop from \Cref{fig:concurrency-bug} as an example,
with {\tt x++;} expanded into its three corresponding pseudo-assembly instructions,
\Cref{fig:tree}(a) shows all possible execution interleavings
%of the compiled code
between 2 threads.

\begin{figure}[p]
	\begin{tabular}{c}
		\includegraphics[width=\textwidth]{statespace-list.pdf}
		\\
		(a) Interleavings visualized individually, as a list.
		\\
		\\
		\includegraphics[width=\textwidth]{statespace-tree.pdf}
		\\
		(b) Interleavings (same order as in (a)), with common prefixes \\
		combined as ``preemption points'', forming a tree.
	\end{tabular}
	\caption[Visualization of interleaving state space for the program in \Cref{fig:concurrency-bug}.]
		{Visualization of interleaving state space for the program in \Cref{fig:concurrency-bug}.
	Thread~1 is represented by purple ovals, thread~2 by yellow squares, and time flows from top to bottom.
	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
	the full state space is twice the size, and the other half is symmetric to the one shown.}
	\label{fig:tree}
\end{figure}

@@

During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
where interrupting the current thread to run a different one is likely to produce different behaviour.
These so-called {\em preemption points} may be identified by any combination of human intuition and machine analysis.
Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
Considering that at each preemption point multiple threads exist as options to run next,
the set of possible ways to execute the program can be viewed as a tree.
\Cref{fig:tree}(b) shows a visualization of the corresponding tree from our example program,
using each pseudo-assembly instruction as a preemption point.

The number of preemption points in each execution defines the depth of this tree,
and the number of threads available to run defines the branching factor.
Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
Addressing the scaling problem in this exponential relation is the central research problem for all model checkers,
\revisionminor{which this thesis will address in \Cref{chap:quicksand}}.

@@

{\bf Concretely speaking:}
\Cref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
The specifics of execution before the thread 1/thread 2 sequence,
other possible threads to run instead of threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
\cref{sec:landslide-dpor} explains how DPOR implements such a deduction in more detail.

@@

Iterative Context Bounding \cite{chess-icb} is a popular such technique which heuristically reorders the search to prioritize interleavings with fewer preemptions first.
This heuristic is based on the insight that most bugs require few preemptions to uncover, so interleavings with a number of preemptions that exceeds a certain bound will be de-prioritized, only tested until after all the fewer-preemption interleavings are completed.
Preemption sealing \cite{sealing} is another heuristic strategy which restricts the scope of the search by limiting the model checker to use only preemption points arising from certain functions in the source code.
This allows developers to vastly reduce state space size by identifying which program modules are already trusted,
although it requires some human intuition to correctly mark those boundaries.
Iterative Deepening, presented in \Cref{chap:quicksand}, is another such search heuristic.

@@

Two instructions are said to race if:
\begin{enumerate}
	\item they both access the same memory address,
	\item at least one is a write,
	\item the threads do not hold the same lock,
	\item and no synchronization enforces an order on the thread transitions
		(the {\em Happens-Before} \revisionminor{relation~\cite{lamport-clocks,hybriddatarace}}, described below).
\end{enumerate}
%In \Cref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
In \Cref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.

@@

\begin{itemize}
	\item Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
as shown in the example program in \Cref{fig:hb-example}(a).
We call such unreported access pairs {\em false negatives}.

\item
\cite{hybriddatarace} introduces the {\em Limited Happens-Before} relation,
which will report such potential races
by considering only blocking operations like {\tt cond\_wait} to enforce the order.
However, consider the similar program in \Cref{fig:hb-example}(b),
in which the access pair ceases to exist in the alternate interleaving.
Limited Happens-Before will report all potential races, avoiding false negatives \cite{tsan},
but at the cost of necessarily reporting some such {\em false positives}.

\item
In recent work, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
extends Happens-Before to additionally report a subset of potential races while soundly avoiding false positives.
It tracks conflicting accesses in intervening
critical sections to determine whether lock events are unrelated to a potential race.
Causally-Precedes will identify the potential race in \Cref{fig:hb-example}(a), as the two critical sections do not conflict,
although it can still miss true potential races in other cases.
\end{itemize}

Landslide implements both Happens-Before (henceforth referred to as {\em Pure Happens-Before} for clarity) and Limited Happens-Before.
\Cref{chap:quicksand} includes a comparison of the two approaches for the purpose of finding new preemption points for model checking.
%, we use the Limited Happens-Before relation for our analysis.
%The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
%my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
%Hence, I accept some overhead from false positives for the sake of more thorough testing.

@@

The Pebbles kernel architecture
is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and \revisionminor{Implementation} \cite{kspec,thrlib}.
In the course of a semester, students work on five programming assignments;
the first two are individual, and the remaining three are the products of two-person teams.
I will focus on the third and fourth of these, the thread library and kernel,
called ``P2'' and ``P3'' respectively (the project numbers start at 0).
The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in \Cref{tab:syscalls}.

@@

\subsubsection{P3}
In P3, students implement a kernel which provides the same system calls shown in \Cref{tab:syscalls}, previously provided by the reference kernel.
Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
This requires less implementation complexity than the more featureful
\revisionminor{{\tt rfork} model, of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model, of Linux}.

@@

The 15-410 course staff is notoriously secretive about the nature of many concurrency bugs
students commonly encounter during P2 and P3.
This is driven by a desire to cause students to find, diagnose, and fix these bugs on their own during the projects,
rather than to be surprised by them afterwards during grading
\cite{de0u-2018}.
%
One such example is the {\tt paraguay} unit test distributed with P2 (\cref{sec:education-pebbles-tests}),
which targets a subtle condition-variable bug.
The test uses the {\tt misbehave} system call to \revisionminor{amplify the probability of} a particular thread interleaving,
likely to expose the bug,
which is otherwise very unlikely to arise in normal execution.
The reference kernel specification \cite{kspec} does not define the {\tt misbehave} modes' behaviours,
as doing so would deprive students of the learning experience of discovering the interleaving in question on their own.
%
I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.

@@

		see \cref{sec:background-datarace}.

@@

		(\cref{sec:quicksand-soundness}).

@@

		See \cref{sec:landslide-dpor}.

@@

		See \cref{sec:landslide-estimate}.

@@

		See \cref{sec:overview-statespace}; \Cref{fig:tree}.

@@

		are both {\em heuristic} (\cref{sec:landslide-blocking}),

@@

	\item {\bf ICB.} Short for ``Iterative Context Bounding''. See \cref{sec:landslide-icb}.

@@

		See \cref{sec:quicksand-id}.

@@

		Documented in \Cref{chap:landslide}.

@@

		See \cref{sec:overview-mutex}.

@@

		is kept secret to preserve the intended challenge of debugging (see \cref{sec:410-secrecy}).}

@@

		See \cref{sec:pebbles}.

@@

		The educational kernel architecture used in 15-410 at CMU. See \cref{sec:pebbles}.}

@@

		as well as several other universities. See \cref{sec:overview-pintos}.}

@@

		See \cref{sec:landslide-pps}.

@@

		See \Cref{chap:quicksand}.
